
---
title: "RMarkdown Building Blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{0203-rmdBuildingBlocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<!-- set root directory here
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/markdownR/markdown_html")
knitr::opts_chunk$set(warning=FALSE, fig.align="center")
```
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Last update: `r format(Sys.time(), '%d %B %Y')` (V2019.2)

---

## Overview

The minimalist RMarkdown document consists of three parts:

* A YAML; and
* Formatted text.

With these first two parts you could write a simple text letter using RMarkdown.  However, this ignores the power of RMarkdown, which in the world of analysis allows us to next:

* Embed R code for execution.

These three elements -- the YAML, formatted text, and executable R code -- constitute the basic RMarkdown document.  Once this document is built, it can be "knit," resulting a document not that not only includes explanatory and interpretative text but also provides a metadocument of all analytical code and output.

## The Basic Structure - The YAML Header

All RMarkdown documents begin with a **YAML header**, which originally meant "Yet Another Markup Language," but has since morphed to mean "YAML Ain't Markup Language."  The YAML header controls the format of the RMarkdown document.  A basic YAML for a HTML document generated from within RStudio looks like the one found in our **hmwk1.rmd** file:

```
#### What a basic YAML looks like
---
title: "Homework 1"
author: "Student R. Me"
date: "August 15, 2019"
output: html_document
---
```

Of the four YAML elements shown above, only the **output:** is truly needed.  This YAML requests output be a **html_document**.  Some possible, but by no means all, RMarkdown output types include:

* **output: pdf_document** (Adobe PDF)
* **output: word_document** (MS WORD)
* **output: beamer_presentation** (Adobe PDF slideshow)
* **output: ioslides_presentation** (HTML slideshow)

A more complete list of RMarkdown formats can be found [here](https://rmarkdown.rstudio.com/formats.html).

Other document control features, such as a CSS file that controls HTML output, or a MS WORD document template to control MS WORD output, can be added to the YAML.  More on that a bit later. 

## The Basic Structure -- The Text

```
####  Nothing special about adding text
This is text.  Type away to your heart's content ...

There's a rudimentary spell--checker, as well as a search icon, in the toolbar of the upper left RStudio window where you compose the RMD file.

You can also bring text in from elsewhere, such as cut--n--paste from a MS WORD document.  If I'm writing lots of text I often write in MS WORD and use the spell--checker there before inserting into the RMD document.
```

Click [here to see what the text output looks like](https://tedwards-github.github.io/useRfiles/rmd-files/rmd_basics01.html).  We'll learn more about how to configure text for bold, italics, and other text formatting features next.

## The Basic Structure -- Horizontal Line Separators

The most common use of horizontal separators is to distinguish among document sections.  You can add one, two, or any number of line separators.  Note a hard return and blank line is required after the first separator to generate two lines !!

```
#### Coding for line separators
The sequence of --- (three hyphens) adds a line separator to the document.

---

The sequence two sets of --- (three hyphens) separated by a blank line

---

---

adds two lines of separators.  
```

## The Basic Structure -- Text Headers

Headers can be up to 6 levels, with each level preceded by the number of `#` (pound or hashtag) symbols representing the header level.

```
#### Adding different levels of headers
# Header1
## Header2
### Header3
#### Header4
##### Header5
###### Header6
```

[Examine the header output](https://tedwards-github.github.io/useRfiles/rmd-files/rmd_basics02.html), showing both headers and line separators.  Note that by the time you've reached **Header6** the font is quite small relative to the text.  You can create headers with greater than 6 `#`; however, at some point headers $>$ 6 revert to the base pitch.

## The Basic Structure -- Bold, Italics, Blockquotes, and Lines As Separators

Words and phrases can be emphasized using **Bold**, _Italics_, and _**Combinations**_ of each.  There's even a **Blockquote** option which allows for (pithy?) phrases or paragraphs to be highlighted.

```
#### Boldface and Italics
Encapsulate a word with ** (two asterik) or __ (two underscores) 
to **Bold** or __Bold__ text. 

For italics, use a * (single asterisk) or _ (single underscore) 
for *emphasis* or _emphasis_. 

Use a sequence of *** (three astrik) or ___ (three underscores) 
to create ***bolded, italicized words***. 

There should be no spaces betwen the first and last characters of the 
word or phrase to be bolded or italicized.

#### The Blockquote
Use the > (greater than) symbol to create a

> Blockquote of (typically pithy) text.
```

Our [RMarkdown document so far](https://tedwards-github.github.io/useRfiles/rmd-files/rmd_basics03.html) includes basic text, line separators and headers to distinguish among document sections, and boldface and italics.  There's even formatting to add blockquotes that highlight text sections.

Let's next add lists.

## The Basic Structure -- The List

Lists use the `*` (asterisk) or `+` (plus) or `-` (hyphen) for an unordered list; simply add a hard return and another `*` or `+` or `-` to create the second item in the list.  For indenting, use either a tab or at least four spaces.

A numbered list begins with the desired number, say 1, and followed with successive numbers.

Note that RMarkdown has some limitations for classical numeric listing of 1, then 1.1, then 1.1.1, etc.  These are better controlled using a CSS template from within the YAML.

```
#### Unordered Lists
* Build an unordered Markdown list using the * (asterisk)
* Add a second list element
    * at least 4 spaces (or a tab) before the * results in indenting
    
#### Ordered Lists
1. Build an ordered Markdown list using numbers
2. Add a second list element
    1. at least 4 spaces (or a tab) before an ordered sub-list
```

Click [here to see what list output looks like](https://tedwards-github.github.io/useRfiles/rmd-files/rmd_basics04.html).

## The Basic Structure -- Embedding R Code

Any R--based coding operations can be embedded into a RMarkdown document and resultant output displayed.  This means if you've crazy complex R code you can run that code from a RMarkdown document and create a report containing all code, its output, plus whatever King's English you wish to include. 

However, at this stage in learning how to use RMarkdown, we'll keep the R coding simple  until we've learned how to build a basic RMarkdown document.  More complex operations -- complete with fully formed examples -- are shown below.

Specifically, we will:

* Learn about the **chunk** for embedding code
* Embed and control execution of R code within a RMarkdown document

### What is the **chunk** in a RMarkdown document?

**chunk** is the term that refers to R code to be rendered in a RMarkdown document.  A **chunk** can include one or more lines of R code implemented as a block of code, or a **chunk** can be executed inline and directly from within text constituting a sentence.  **chunk**'s can be embedded directly by typing into the document, or by using the drop-down **chunk** icon on the toolbar. 

The basic block **chunk** begins with ` ```{r}` (three backticks ` ``` ` then `{r}` with no spacing), follows with a hard returns for each line of code in the block, and ends with ` ``` ` (three backticks).

An inline **chunk** of code begins with `` `r ``, follows with a single blank space, then the executable R code `Code2Execute`, and ends with `` ` ``.  

    #### The basic R Markdown chunk as a block of code 
    ```{r}`r ''`
    CODE TO EXECUTE GOES HERE
    ```
    #### The basic R Markdown chunk as inline, executed R code
    blah blah text `r`r ''` Code2Execute` more blah blah text

The default execution of a block **chunk** returns the R code plus results of the code, while an inline **chunk** returns a value and embeds it directly into a sentence.

### Setting local chunk options

Common options to modify output to the RMarkdown document include: 

* Naming a **chunk**, which is useful when searching for errors;
* Return results but not the code; 
* Return the code, or some part of the code, but not the results; 
* Having all or some portion of the R code echoed back during execution.  

These options can be set locally within a **chunk**, or set globally for all **chunk**'s.  Within a **chunk** options appear inline within the `{r}` syntax.  Multiple options can be set, with each separated by a `,`.  Although RStudio shows some of the [available **chunk** options](http://rmarkdown.rstudio.com/authoring_rcodechunks.html), a more comprehensive discussion of possible **chunk** options is found within the [**knitr** homepage](http://yihui.name/knitr/options/).

    #### Naming a chunk
    ```{r ChunkName}`r ''`
    CODE TO EXECUTE GOES HERE
    ```
    #### Return the results but not the code
    ```{r echo=F}`r ''`
    CODE TO EXECUTE GOES HERE
    ````
    #### Return the results, and only lines 4:7 of the code
    ```{r echo=4:7}`r ''`
    CODE TO EXECUTE GOES HERE
    ```
    #### Return the code but do not evaluate
    ```{r eval=F}`r ''`
    CODE TO EXECUTE GOES HERE
    ```
    #### Naming a chunk, and supressing return of code
    ```{r ChunkName, echo=F}`r ''`
    CODE TO EXECUTE GOES HERE
    ```

### Setting global chunk options

Global options set **chunk** options throughout the entire document, such as the suppression of R messages and warnings.  Global options should be inserted after the YAML, and before the document text begins. 

Two common global options are shown below: the first suppresses (the sometimes truly obnoxious R) warnings and messages, while the second sets a global working directory for the RMarkdown document.  **NOTE** that whatever `root.dir` is set to requires all subsequent operations, e.g., access to data files to input, any output directories, be nested underneath that directory.

    #### Setting global options 
    ```{r global options}`r ''`
    knitr::opts_chunk$set(warning=FALSE, message=FALSE)
    knitr::opts_knit$set(root.dir = "~/MyRMarkdownDocumentDirectory")
    ```
You can change to a different directory within a chunk to access data at locations other than your specified default. However, setting a path different from the default works **only** in the chunk where the path is changed.  Once you leave that chunk, the directory defaults to the one initially specified as shown above.

Check out what a [RMarkdown document with embedded code](https://tedwards-github.github.io/useRfiles/rmd-files/rmd_basics05.html) looks like.

