
---
title: "EXAMPLE: A Finalized Document"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{0204-rmdFinalExample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<!-- set root directory here
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/markdownR/markdown_html")
knitr::opts_chunk$set(warning=FALSE, fig.align="center")
```
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

Last update: `r format(Sys.time(), '%d %B %Y')` (V2019.2)

---

## What You Can Expect !!

Assume you have been instructed to complete a formal analysis and write--up of prey capture by male and female osprey using the data from above.  The write--up would include a list of objectives, the R code used for analysis, the output, and some measure of interpretation.  

One possible rendition of this homework assignment in HTML format is [here](https://maddoor1.nr.usu.edu:45231/sharing/cJ2gXlNlH).

You're done!  You should now be able build a RMarkdown document that integrates R coding and analysis directly into a HTML document.

P.S. The cheatsheet *.Rmd file that built this [can be accesssed from here]().
