---
title: "Course--Required R Packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{InstallPackages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installing Course-required Packages

This section is applicable to all workshop participants. Following the procedures below will install all the required packages and dependencies needed for the course.

* Open RStudio or Vanilla R.
* Copy the code for each STEP into a new R Script file (name it `InstallMyPackages.R`, that way you'll have a permanent R script), highlight the code by STEP, and run each STEP.

### Step #1

This Step creates a simple R function called `bulkpak` that loads course--required packages.

```{r, eval=F}
# build install fxn
bulkpak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, repos = "http://cran.r-project.org", dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
```

There are a gazillion ways to install packages; this merely reflects one of those gazillion. Feel free to install each using the time--honored `install.packages()` function, or the RStudio--based Install GUI, if you prefer.

### Step #2

This STEP includes the bare minimum packages for **sdhmR-V2019.2**.  The `bulkpak` function above merely calls the packages in the object `packages` and installs them unless they already exist in your R library.

```{r, eval=F}
# call function bulkpak and install these listed packages
packs_sdhmR <- c("raster","rgdal", "rgeos", "maptools", "sp", "GISTools", 
                 "PresenceAbsence", "DAAG", "gam", "dismo", "randomForest", "gbm", 
                 "spocc", "sf", "sampling")
bulkpak(packs_sdhmR)
```

### Step #3

If you prefer to work in a `tidyverse` rather than a base R environment, then you must install (or ensure are already installed) the following packages as well.  Note that although packages `readxl` and `lubridate` work with the `tidyverse` environment they must be installed separately.  

For reasons beyond this mere mortal's understanding, the `tidyverse` environment does not export back to **.xlsx**/**.xls** formats.  You must instead use package `xlsx`.

```{r, eval=F}
# call function bulkpak and install these listed packages
packs_tidy <- c("tidyverse","readxl", "lubridate", "xlsx")
bulkpak(packs_tidy)
```

You're done.  Installation of Java is next.
