---
title: "Install MAXENT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{InstallMAXENT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Obtain and Install MAXENT

To obtain MAXENT go to:

```{r,eval=F}
http://rob.schapire.net/
```

Click on MAXENT software and datasets.

Register and download the default current version (3.4.1, as of 21 May 2019) into your default (or wherever) downloads directory.  Unzip the file.  Once unzipped you will see four files:  

* maxent.bat, 

* maxent.jar, 

* maxent.sh, and 

* readme.txt

## Install Package `dismo`

Next, install package `dismo` and dependencies in R, if you have not already done so.  `dismo` is the R interface for MAXENT (as well as other SDM tools).

```{r, eval=F}
# install dismo if not installed using scripts in Install Packages
install.packages("dismo", dependencies = T)
```

`dismo` will (if you have a standard configuration) be in your active win--library, and will have a directory called `java`.

Copy and paste, or move, the `maxent.jar`, `maxent.bat`, `maxent.sh`, and `readme.txt` files into the `java` directory.

MAXENT can now accessed from any R session once package dismo is loaded.  We will see how to access MAXENT during the class.

Problems?

Start by reading the `readme.txt`.  Most problems can be solved from there.

You're done.




